---
title: "KEGG analysis"
author: "Ha M. Tran"
date: "2023-12-07"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  eval = TRUE,
  fig.align = "center"
)
```

```{r load libraries}
# working with data
library(dplyr)
library(magrittr)
library(readr)
library(tibble)
library(reshape2)
library(tidyverse)
library(KEGGREST)
library(data.table)

# Visualisation:
library(kableExtra)
library(ggplot2)
library(grid)
library(pander)
library(viridis)
library(cowplot)
library(pheatmap)
library(DT)
library(extrafont)

# Custom ggplot
library(ggplotify)
library(ggpubr)
library(ggbiplot)
library(ggrepel)

# Bioconductor packages:
library(edgeR)
library(limma)
library(Glimma)
library(clusterProfiler)
library(org.Mm.eg.db)
library(enrichplot)
library(pathview)

library(pandoc)
library(knitr)
opts_knit$set(progress = FALSE, verbose = FALSE)
opts_chunk$set(warning=FALSE, message=FALSE, echo=FALSE)
```

```{r importData}
# load DGElist previously created in the set up
dge <- readRDS(here::here("0_data/rds_objects/dge.rds"))
fc <- readRDS(here::here("0_data/rds_objects/fc.rds"))
fdr <- readRDS(here::here("0_data/rds_objects/fdr.rds"))
lm <- readRDS(here::here("0_data/rds_objects/lm.rds"))
lm_sig <- readRDS(here::here("0_data/rds_objects/lm_sig.rds"))

# to increase the knitting speed. change to T to save all plots
savePlots <- F
export <- F
```

```{r importFunctions}
# Theme
bossTheme <- readRDS(here::here("0_data/functions/bossTheme.rds"))
bossTheme_bar <- readRDS(here::here("0_data/functions/bossTheme.rds"))
groupColour <- readRDS(here::here("0_data/functions/groupColour.rds"))
groupColour_dark <- readRDS(here::here("0_data/functions/groupColour_dark.rds"))
expressionCol <- readRDS(here::here("0_data/functions/expressionCol.rds"))
expressionCol_dark <- readRDS(here::here("0_data/functions/expressionCol_dark.rds"))


DT <- readRDS(here::here("0_data/functions/DT.rds"))

# Plotting
convert_to_superscript <- readRDS(here::here("0_data/functions/convert_to_superscript.rds"))
exponent <- readRDS(here::here("0_data/functions/exponent.rds"))
format_y_axis <- readRDS(here::here("0_data/functions/format_y_axis.rds"))
```

# KEGG Analysis

*KEGG pathway images reproduced by permission from Kanehisa Laboratories, September 2023*

KEGG analysis, or Kyoto Encyclopedia of Genes and Genomes analysis, is a method that involves the mapping of molecular datasets, such as DE genes or proteins, to the reference pathways in the KEGG database. The KEGG database provides a comprehensive resource for understanding the molecular interaction and reaction networks within biological systems. KEGG pathways encompass a wide range of biological processes, including metabolism, cellular processes, environmental information processing, and human diseases.

<details>

<summary>More information about KEGG</summary>

The KEGG database is organized into several classes:

-   **Pathway Maps (PATH):** This class includes diagrams of molecular interactions and reactions in various biological pathways. Each pathway map is associated with specific biological processes or functions.

-   **BRITE (B):** BRITE is a hierarchical classification of biological entities, such as genes, proteins, and compounds. It provides a functional hierarchy and relationships between different biological components.

-   **Module (M):** Modules are sets of manually defined functional units, which can represent functional modules of genes or proteins in specific pathways.

-   **Orthology (KO):** The Orthology class provides information about orthologous gene groups, which are genes in different species that evolved from a common ancestral gene. This class is particularly useful for comparative genomics.

For this analysis, only KEGG pathway database will be used. This database is further sub-categorised into several classes:

1.  Metabolism
2.  Genetic Information Processing
3.  Environmental Information Processing
4.  Cellular Processes
5.  Organismal Systems
    1.  Immune system (i.e. T cell receptor signaling pathway , Th1 and Th2 cell differentiation & etc.)
    2.  Endocrine system (i.e. Estrogen signaling pathway , Progesterone-mediated oocyte maturation & etc.)
6.  Human Diseases
7.  Drug Development

</details>

```{r enrichKEGG, eval=FALSE}
enrichKEGG <- list()
enrichKEGG_all <- list()
enrichKEGG_sig <- list()
for (i in 1:length(fc)) {
  for (j in 1:length(fdr)) {
    name <- paste0(fc[i] %>% as.character(), "_", fdr[j] %>% as.character())
    
    # find enriched KEGG pathways
    enrichKEGG[[name]] <- clusterProfiler::enrichKEGG(
      gene = lm_sig[[name]]$entrezid,
      keyType = "kegg",
      organism = "mmu",
      pvalueCutoff = 0.05,
      pAdjustMethod = "none",
      qvalueCutoff = 1
    )
  }
}


for (i in 1:length(fc)) {
  for (j in 1:length(fdr)) {
    name <- paste0(fc[i] %>% as.character(), "_", fdr[j] %>% as.character())
    
    enrichKEGG[[name]] <- enrichKEGG[[name]] %>% 
      clusterProfiler::setReadable(OrgDb = org.Mm.eg.db, keyType = "ENTREZID")
    
    
    enrichKEGG[[name]]@result$Description <- sub(pattern = " \\- Mus musculus \\(house mouse\\)",
                                                 replacement = "",
                                                 x = enrichKEGG[[name]]@result$Description)
    
    # at the beginnning of a word (after 45 characters), add a newline. shorten the y axis for dot plot
    enrichKEGG[[name]]@result$Description <- sub(pattern = "(.{1,45})(?:$| )",
                                                 replacement = "\\1\n",
                                                 x = enrichKEGG[[name]]@result$Description)
    
    # remove the additional newline at the end of the string
    enrichKEGG[[name]]@result$Description <- sub(pattern = "\n$",
                                                 replacement = "",
                                                 x = enrichKEGG[[name]]@result$Description)
    
    enrichKEGG[[name]]@result$Description <- enrichKEGG[[name]]@result$Description %>% str_to_sentence()
    
    
    enrichKEGG_all[[name]] <- enrichKEGG[[name]]@result
    
    # filter the significant and print top 30
    enrichKEGG_sig[[name]] <- enrichKEGG_all[[name]] %>%
      dplyr::filter(pvalue <= 0.05) %>%
      separate(col = BgRatio, sep = "/", into = c("Total", "Universe")) %>%
      dplyr::mutate(
        logPval = -log(pvalue, 10),
        GeneRatio = Count / as.numeric(Total)
      ) %>%
      dplyr::select(c("Description", "GeneRatio", "pvalue", "logPval", "p.adjust", "qvalue", "geneID", "Count"))
    
    
  }
}
```

```{r keggDetails, eval=FALSE}
uniq_id <- lapply(enrichKEGG_all, function(x) rownames(x) %>% as.data.frame)
uniq_id <- data.table::rbindlist(uniq_id) %>% unique()
id_groups <- split(uniq_id$., (seq_along(uniq_id$.) - 1) %/% 10 + 1)

pathways <- flatten(map(id_groups, ~keggGet(.x)))


lapply(id_groups, function(ids) keggGet(ids))
  
  c(keggGet(uniq_id$.[1:10]),
              keggGet(uniq_id$.[11:20]),
              keggGet(uniq_id$.[21:29]))

# Extract relevant information and create a data frame
pathways_details <- map_df(pathways, ~{
  kegg_id <- .$ENTRY
  data.frame(
    "kegg.id" = kegg_id,
    "path" = .$PATHWAY_MAP,
    "class" = if ("CLASS" %in% names(.)) .$CLASS else NA,
    "path.des" = if ("DESCRIPTION" %in% names(.)) toString(.$DESCRIPTION) else NA,
    "rel.path" = if ("REL_PATHWAY" %in% names(.)) toString(.$REL_PATHWAY) else NA,
    "ref" = if ("REFERENCE" %in% names(.)) {
      ref <- do.call("rbind", .$REFERENCE) %>% as.data.frame()
      toString(ref$REFERENCE)
    } else NA
  )
}) 

pathways_details$ref <- gsub("PMID:","",pathways_details$ref)
colnames(pathways_details) <- c("KEGG ID","Pathway","Class","Pathway description","Related pathways","References (PMID)")

saveRDS(pathways_details, here::here("0_data/rds_objects/pathway_details.rds"))
```


## Visualisation {.tabset .tabset-pills}

For data exploratory purposes, the following visualisations are KEGG enrichment analysis performed with set of DE genes significantly below **FDR \< 0.1 or \< 0.05** with and without FC threshold (TREAT). **IMPORTANTLY**, these KEGG terms are all significantly enriched but only with P-value \< 0.05 and no P-value correction method.

-   **Dot plot:** illustrates the top 7 enriched KEGG pathways

    -   $Gene ratio =$ the number of significant DE gene in the term / the total of number of genes in the term as indicated by the size

-   **Table:** list of all the significant KEGG pathways

    -   NOTE: To keep this a readable table, the full pathway description were removed, check the exported Excel spreadsheet for full details on pathways class, descriptions, related pathways, and references

-   **Upset:** illustrate the overlap of gene between different pathways

**I recommend reading through the full list of significant KEGG pathways and selecting the most biologically relevant for better visualisation**

### FC=none, FDR\<0.1 {.tabset .tabset-pills}

#### Dot plot

```{r kegg_dot}
enrichKEGG <- readRDS(here::here("0_data/rds_objects/enrichKEGG.rds"))
enrichKEGG_all <- readRDS(here::here("0_data/rds_objects/enrichKEGG_all.rds"))
enrichKEGG_sig <- readRDS(here::here("0_data/rds_objects/enrichKEGG_sig.rds"))

kegg_dot <- list()
kegg_tab <- list()
kegg_upset <- list()
for (i in 1:length(fc)) {
  for (j in 1:length(fdr)) {
    name <- paste0(fc[i] %>% as.character(), "_", fdr[j] %>% as.character())    
    
    # dot plot, save
    kegg_dot[[name]] <- ggplot(enrichKEGG_sig[[name]][1:7, ]) +
      geom_point(aes(x = GeneRatio, y = reorder(Description, GeneRatio), colour = logPval, size = Count)) +
      scale_color_gradientn(colors = rev(c("#FB8072","#FDB462","#ffffd5","#8DD3C7","#80B1D3")),
                            values = scales::rescale(c(min(enrichKEGG_sig[[name]]$logPval), max(enrichKEGG_sig[[name]]$logPval))),
                            breaks = scales::pretty_breaks(n = 5)) +
      scale_size(range = c(2,5)) +
      labs(x = "Gene ratio", y = "", color = expression("-log"[10] * "P value"), size = "Gene Counts") +
      bossTheme(base_size = 14,legend = "right")

    kegg_tab[[name]] <- enrichKEGG_sig[[name]] %>% 
      dplyr::mutate_if(is.numeric, funs(as.character(signif(.,3)))) %>% 
      DT(., caption = "Significantly enriched KEGG terms")
    
    
    kegg_upset[[name]] <- upsetplot(x = enrichKEGG[[name]], 7)
    
    if(savePlots == TRUE) {
      ggsave(filename = paste0("kegg_dot_", name, ".svg"), plot = kegg_dot[[name]], path = here::here("2_plots/3_FA/kegg/"),
             width = 18, height = 16, units = "cm")
      ggsave(filename = paste0("kegg_upset_", name, ".svg"), plot = kegg_upset[[name]], path = here::here("2_plots/3_FA/kegg/"), 
             width = 20, height = 12, units = "cm")
      
    }
    
  }
}

kegg_dot[[1]]
```

#### Table

```{r tab}
kegg_tab[[1]]
```

#### Upset Plot

```{r upset}
kegg_upset[[1]]
```

```{r loop, results="asis"}
## this function is basically creating chunks within chunks, and then
## I use results='asis' so that the html image code is rendered 
kexpand <- function(wd, ht, cap) {
  cat(knit(text = knit_expand(text = 
     sprintf("```{r %s, fig.width=%s, fig.height=%s}\n.pl\n```", cap,wd, ht)
)))}


# Loop through each FC value
headers <- c("FC=none, FDR<0.1", "FC=none, FDR<0.05", "FC=1.1, FDR<0.1", 
             "FC=1.1, FDR<0.05", "FC=1.2, FDR<0.1"  , "FC=1.2, FDR<0.05")
types <- c("Dot plot", "Table", "Upset plot")

for (i in 2:length(headers)) {
  cat(paste0("### ",headers[i],"{.tabset .tabset-pills} \n\n"))
  
  cat(paste0("#### ",types[[1]]," \n"))
  .pl <- kegg_dot[[i]] 
  kexpand(wd = 8,ht = 8,cap = paste0("dot",i))
  cat("\n\n")
  
  cat(paste0("#### ",types[[2]]," \n"))
  .pl <- kegg_tab[[i]]
  kexpand(wd = 11,ht = 8,cap = paste0("tab",i))
  cat("\n\n")
  
  cat(paste0("#### ",types[[3]]," \n"))
  .pl <- kegg_upset[[i]] 
  kexpand(wd = 11,ht = 9,cap = paste0("upset",i))
  cat("\n\n")
}
```

# Pathway specific heatmaps {.tabset .tabset-pills}

```{r}
# chosing the pathways of interest
kegg_id <- c("mmu04610", "mmu04080", "mmu04310")

# t <- sapply(enrichKEGG_sig, function(x) rownames(x)) 
kegg_pathway <- KEGGREST::keggGet(kegg_id)


```

Here the top `r length(kegg_id)` most interesting KEGG pathways were selected for further visualisations.

These visualisations are similarly performed with set of DE genes significantly below **FDR \< 0.1 or \< 0.05** with and without FC threshold (TREAT). **IMPORTANTLY**, these KEGG terms are all significantly enriched but only with P-value \< 0.05 and no P-value correction method.

-   **Heatmap:** illustrates the expression of genes in specific KEGG pathways.

    -   The genes (rows) and samples (columns) are heirachirally clusterd based on their expression profile.

-   **Table:** list of all the significant DE genes in the specified KEGG pathway

-   **Pathview:** maps gene expression data onto the specified KEGG pathways, allowing users to see where genes in their dataset are located within specific pathways, and potential upstream and downstream elements.

```{r heatmap}
# create df with normalised read counts with an additional entrezid column for binding
logCPM <- cpm(dge, prior.count = 4, log = TRUE) %>% subset(select = 1:8)
rownames(logCPM) <- dge$genes$gene

# df for heatmap annotation of sample group
anno <- as.factor(dge$samples$group) %>% as.data.frame() %>% dplyr::slice(1:8)
colnames(anno) <- "Groups"
rownames(anno) <- colnames(logCPM)

matrix <- list()
display_matrix <- list()
kegg_heat=list()

for (i in 1:length(lm)) {
  name <- names(lm[i])
  
  for (j in 1:length(kegg_id)) {
    pathName <- kegg_pathway[[j]]$PATHWAY_MAP
    
    partial <- enrichKEGG_all[[name]][, c("ID", "geneID")]
    partial <- partial[kegg_id[j], "geneID"] %>% as.data.frame()
    partial <- separate_rows(partial, ., sep = "/")
    colnames(partial) <- "entrezid"

    # heatmap matrix
    match <- rownames(logCPM) %in% partial$entrezid
    matrix[[name]][[pathName]] <- logCPM[match, ] %>% as.matrix()
    
    # display matrix
    match2 <- lm_sig[[name]][, "gene"] %in% partial$entrezid
    display_matrix[[name]][[pathName]] <- lm_sig[[name]][match2, c("gene", "logFC", "P.Value", "adj.P.Val", "description","entrezid")] %>%
      dplyr::mutate_if(is.numeric, funs(as.character(signif(.,3)))) %>%
      DT(.,caption = "DE genes")
    
    kegg_heat[[name]][[pathName]] <- ComplexHeatmap::pheatmap(
      # MAIN
      mat = matrix[[name]][[pathName]],
      color = colorRampPalette(rev(c("#FB8072","#FDB462","#ffffd5","#8DD3C7","#80B1D3")))(100),
      scale = "row",
      
      # Col
      cluster_cols = T,
      cutree_cols = 2,
      border_color = "white",
      # gaps_col = 4,
      angle_col = "0",
      treeheight_col = 20,
      
      # Row
      cutree_rows = 4,
      treeheight_row = 30,
      clustering_distance_rows = "euclidean",
      labels_row = as.expression(lapply(rownames(matrix[[name]][[pathName]]), function(a) bquote(italic(.(a))))),
      
      # Labs
      main = paste0(pathName, "\n"),
      show_colnames = T,
      legend = T,
      heatmap_legend_param = list(title = "Z-score", legend_width = unit(3, "cm")),
      
      # Annotation
      annotation_legend = T,
      # legend_labels = T,
      annotation_col = anno,
      annotation_colors = list("Groups" = groupColour),
      annotation_names_col = T,
      
      # Fonts
      fontfamily = "Arial Narrow",
      fontsize = 14,
      fontsize_col = 14,
      fontsize_number = 14,
      fontsize_row = 14
      
    ) %>% as.ggplot()
    
    if (savePlots == T) {
      ggsave(filename = paste0("heat_", name, "_", pathName, ".svg"), 
           plot = kegg_heat[[name]][[pathName]], 
           path = here::here("2_plots/3_FA/kegg/"),
           width = 12,
           height = 14,
           units = "cm")
    }
  }
}
```

```{r pathview, eval=FALSE}
# adjusting the kegg id to suit the parameters of the pathview funtion
adj.keggID <- gsub("mmu", "", kegg_id)

for (i in 1:length(lm)) {
  name <- names(lm[i])
  
  # extract the logFC from the DE gene list
  pathview_table <- dplyr::select(.data = lm_sig[[name]], c("logFC")) %>% as.matrix()
  # run pathview with Ensembl ID instead of entrezID
  pathview <- pathview(
    gene.data = pathview_table[, 1],
    gene.idtype = "SYMBOL",
    pathway.id = adj.keggID,
    species = "mmu",
    out.suffix = "pv",
    kegg.dir = here::here("2_plots/3_FA/"),
    kegg.native = T
  )

  # move the result file to the plot directory
  file.rename(
    from = paste0("mmu", adj.keggID, ".pv.png"),
    to = here::here(paste0("2_plots/3_FA/kegg/pv_", name, "_", kegg_id, ".png"))
  )
}
```


```{r loop2, results="asis"}
## this function is basically creating chunks within chunks, and then
## I use results='asis' so that the html image code is rendered 
kexpand <- function(wd, ht, cap) {
  cat(knit(text = knit_expand(text = sprintf("```{r %s, fig.width=%s, fig.height=%s}\n.pl\n```", cap,wd, ht)
  )))
  }


# Loop through each FC value
headers <- c("FC=none, FDR<0.1", "FC=none, FDR<0.05", "FC=1.1, FDR<0.1", 
             "FC=1.1, FDR<0.05", "FC=1.2, FDR<0.1"  , "FC=1.2, FDR<0.05")
# subheaders <- 
types <- c("Heatmap", "Table", "Pathview")

for (i in 1:length(headers)) {
  cat(paste0("## ",headers[i],"{.tabset .tabset-pills} \n\n"))
  
  for (j in 1:length(kegg_pathway)) {
    cat(paste0("### ",kegg_pathway[[j]]$PATHWAY_MAP,"{.tabset .tabset-pills} \n\n"))
    
    cat(paste0("#### ",types[[1]]," \n"))
    .pl <- kegg_heat[[i]][[j]] 
    kexpand(wd = 8,ht = 9,cap = paste0("heat",i,j))
    cat("\n\n")
    
    cat(paste0("#### ",types[[2]]," \n"))
    .pl <- display_matrix[[i]][[j]]
    kexpand(wd = 11,ht = 8,cap = paste0("specficTab",i,j))
    cat("\n\n")
    
    cat(paste0("#### ",types[[3]]," \n"))
    cat(paste0("![](figure/kegg.Rmd/pv_", names(kegg_heat)[i], "_", kegg_id[j], ".png)"))
    # .pl <- knitr::include_graphics(here::here(paste0("figure/kegg.Rmd/pv_", names(kegg_heat)[i], "_", kegg_id[j], ".png")))
    # kexpand(wd = 9,ht = 9,cap = paste0("pathview",i,j))
    cat("\n\n")
    
  }
}

```

# Export Data

The following are exported:

-   **KEGG_all.xlsx** - This spreadsheet contains all KEGG pathways

-   **KEGG_sig.xlsx** - This spreadsheet contains all significant (P value < 0.05) KEGG pathways

```{r exportData, eval=export}
pathways_details <- readRDS(here::here("0_data/rds_objects/pathway_details.rds"))

# save to csv
for (i in 1:length(enrichKEGG)){
  name <- names(enrichKEGG[i])
  
  enrichKEGG_all[[name]] <- cbind(enrichKEGG_all[[name]],pathways_details[pathways_details$`KEGG ID` %in% rownames(enrichKEGG_all[[name]]),3:6])
  
  enrichKEGG_sig[[name]] <- cbind(enrichKEGG_sig[[name]],pathways_details[pathways_details$`KEGG ID` %in% rownames(enrichKEGG_sig[[name]]),3:6]) 
}

writexl::write_xlsx(x = enrichKEGG_all, here::here("3_output/KEGG_all.xlsx"))
writexl::write_xlsx(x = enrichKEGG_sig, here::here("3_output/KEGG_sig.xlsx"))

saveRDS(enrichKEGG, here::here("0_data/rds_objects/enrichKEGG.rds"))
saveRDS(enrichKEGG_all, here::here("0_data/rds_objects/enrichKEGG_all.rds"))
saveRDS(enrichKEGG_sig, here::here("0_data/rds_objects/enrichKEGG_sig.rds"))
```
